---
title: "afternoon-practice"
author: "Jake Eisaguirre"
date: "8/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
```

```{r}

view(us_rent_income)
summary(us_rent_income)
unique(us_rent_income)

```


## make wide
```{r}

us_rent_wide <- us_rent_income %>% 
  select(-moe) %>% 
  pivot_wider(names_from = variable,
              values_from = estimate)


```

## practice some tidyverse functions for data wrangling
- filter
- select
- mutate
- case_when
- group %>% summarize
- piping things together in sequence

## dplyr::filter() to keep or exclude rows based on our conditions 

1. exact match (single condition)
only keep observations where the island is Biscoe
```{r}

ex_1 <- filter(.data = penguins, island == "Biscoe")

unique(ex_1$island)

```

2. "and" statement within filter

only keep observations if the penguin is chinstrap and year is 2008
```{r}

ex_2 <- penguins %>% 
  filter(species == "Chinstrap", year == 2008)
  

```

3. Exlude gentoos

keep observations if the penguin specis is not "gentoo"

```{r}

ex_3 <- penguins %>% 
  filter(species != "Gentoo")

```

exercise:

create a subset from penguins the only contains observations for Adelie penguins with a flipper length greater than or equal to 185mm

```{r}

sub_set <- penguins %>% 
  filter(species == "Adelie", flipper_length_mm >= 185)

```


4. the %in% operator

use the %in% operator to look fo *any* matches within a vector, and return T if the  value does exist for our row

we wamt to keep observations where the island is Dream or Torgersen

```{r}

ex_4 <- penguins %>% 
  filter(island %in% c("Dream", "Torgersen"))

# == would not work here because it was subsequently search for it. need to use %in% for multiple things to look for in the same column

# filter() is used for rows
```

## select() to exlude Columns

5. keep a single column by name

only want to keep the species column
```{r}

ex_5 <- penguins %>% 
  select(species)

# re write with out the pipe operator 

ex_5.1 <- select(.data = penguins, species)

```

6. keep multiple columns by name

create a subset that contains species, bill length through flipper length and year 

```{r}

ex_6 <- penguins %>% 
  select(species, bill_length_mm:flipper_length_mm, year)

```

7. choose or exlude columns based on position

```{r}

ex_7 <- penguins %>% 
  select(-4)

```

8. a pipe sequence 

where you first only keep columns for species, body mass, and island. second, filter to only include observations if body mass is greater then 3600

```{r}

ex_8 <- penguins %>% 
  
  select(species, body_mass_g, island) %>% 
  
  filter(body_mass_g >= 3600)

```

## muate() to add or update existing columns, while keeping the ones already there

9. Add a column that has body mass converted to kg

```{r}

ex_9 <- penguins %>% 
  
  mutate(body_mass_kg = body_mass_g / 1000) %>% 
  
  mutate(flipper_length_m = flipper_length_mm / 1000)

```

10. Mutate to update an existing column

any where the string "sen" appears in island replace with "yay"

```{r}

ex_10 <- penguins %>% 
  mutate(island = str_replace(island, pattern = "sen", replacement = "yay"))

```

11. update the class of the bill depth to character

```{r}

ex_11 <- penguins %>% 
  mutate(bill_depth_mm = as.character(bill_depth_mm))

```

## case_when(): your favorite if-else

12. create a new column that contains "small" if penguin flipper lenght is < 185mm,  "medium" otherwise 
```{r}

ex_12 <- penguins %>% 
  mutate(binned_size = case_when(
    
    flipper_length_mm < 185 ~ "small",
    flipper_length_mm >= 185 ~ "medium"
    
  ))

```

Task: create a new column "bird_bill_bin" that contains:

- if bill lenth < 38mm : "short bil"
- if bill length >= 38mm & < 43mm: "medium bill
- if bill length >= 43: "long bill

```{r}

ex_13 <- penguins %>% 
  mutate(bird_bill_bin <- case_when(
    bill_length_mm < 38 ~ "short bill",
    bill_length_mm >= 38 & bill_length_mm < 43 ~ "medium bill",
    bill_length_mm >= 43 ~ "long bill"
      ))

```

## group and summarize to find summary values by group, and return them in a nice summary table

14. group_by to say there are groups within the species column

```{r}

ex_14 <- penguins %>% 
  group_by(species, island, year, sex) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = T),
            sd_flipper_length = sd(flipper_length_mm, na.rm = T))

```

## practie task

singe pipe sequence:

only keep penguins that have a body mass > 3500
only keep columns for species, flipper_length_mm, and sex
create a summary table that contains the mean, sd, max, and min value of flipper length by species

```{r}

ex_15 <- penguins %>% 
  
  filter(body_mass_g > 3500) %>% 
  
  select(species, flipper_length_mm, sex) %>% 
  
  group_by(species, sex) %>% 
  
  summarize(mean_flipper_length_mm = mean(flipper_length_mm, na.rm = T), 
            
            sd_flipper_length_mm = sd(flipper_length_mm, na.rm = T), 
            
            max_flipper_length_mm = max(flipper_length_mm, na.rm = T), 
            
            min_flipper_length_mm = min(flipper_length_mm, na.rm = T))

```




